Relevance-based Quantization of Scattering Features for Unsupervised Mining of Environmental Audio

### 2018-09-12 paper/proof/13636_2018_138_MOESM1.tex
Gammatone wavelets have center freq \xi, not 1
bugfix Fourier transform of Gammatone wavelet
state that \xi and \sigma are expressed in radians
1 -> \xi in text

The asymmetric amplitude profile of Gammatones makes them suitable to model temporal masking in auditory filterbanks~\cite{Fastl2007}.
Yet, the introductory paper on Gammatone wavelets \cite{Venkitaraman2014} does not provide a formula for deducing $\sigma$ from the specification of a quality factor $Q$.
In this appendix, we provide a rationale for choosing the topmost center frequency $\xi$ of a Gammatone wavelet filter bank in a discrete-time setting. Then, we relate the bandwidth parameter $\sigma$ to the choice of a quality factor Q.

@@

The complex-valued Gammatone wavelet is a modification of the real-valued
Gammatone auditory filter, originated in auditory physiology. The
Gammatone auditory filter of dimensionless frequency $1$ is defined
as a gamma distribution of order $N\in\mathbb{N}^{*}$ and bandwidth
$\sigma$ modulated by a sine wave, that is,
\[
t^{N-1}\exp(-2\pi\sigma t)\cos(2\pi t).
\]
For a fixed $\sigma$, the integer $N$ controls the relative shape
of the envelope, becoming less skewed as $N$ increases. Psychoacoustical
experiments have shown that, for $N=4$, the Gammatone function provides
a valid approximation of the basilar membrane response in the mammalian
cochlea \cite{Flanagan1960,Patterson1976,Lyon2010}. In particular,
it is asymmetric both in the time domain and in the Fourier domain,
which allows to reproduce the asymmetry of temporal masking as well
as the asymmetry of spectral masking \cite{Fastl2007}. It is thus
used in computational models for auditory physiology \cite{Pressnitzer2005}.
However, it does not comply with the Grossman-Morlet admissibility
condition, because it has a non-negligible average. In addition, because
the Gammatone auditory filter takes real values in the time domain,
its Fourier transform satisfies Hermitian symmetry, which implies
that it does not belong to the space $H^{2}$ of analytic functions.
More generally, there are no real-valued functions in $H^{2}$ \cite{Grossmann1984}.

@@

With the aim of building a pseudo-analytic admissible Gammatone wavelet,
\cite{Venkitaraman2014} have modified the definition of the Gammatone
auditory filter, by replacing the real-valued sine wave $\cos(2\pi t)$
by its analytic part $\exp(2\pi\mathrm{\mathrm{i}}t)$ and by taking
the first derivative of the gamma distribution, thus ensuring null
mean. The definition of the Gammatone wavelet becomes
\[
\boldsymbol{\psi}(t)=\left(2\pi(\mathrm{i}-\sigma)t^{N-1}+(N-1)t^{N-2}\right)\exp(-2\pi\sigma t)\exp(2\pi\mathrm{i}t)
\]
in the time domain, and
\[
\boldsymbol{\widehat{\psi}}(\omega)=\dfrac{\mathrm{i}\omega\times(N-1)!}{(\sigma+\mathrm{i}(\omega-\sigma))^{N}}
\]
in the Fourier domain. Besides its biological plausibility, the Gammatone
wavelet enjoys a near-optimal time-frequency localization with respect
to the Heisenberg uncertainty principle. Furthermore, this time-frequency
localization tends to optimality as $N$ approaches infinity, because
the limit $N\rightarrow+\infty$ yields a Gabor wavelet \cite{Cohen1995}.
Last but not least, the Gammatone wavelet transform of finite order
$N$ is causal, as opposed to the Morlet wavelet transform, which
makes it better suited to real-time applications. From an evolutionary
point of view, it has been argued that the Gammatone reaches a practical
compromise between time-frequency localization and causality constraints \cite{Venkitaraman2014}.

>>>>

The asymmetric amplitude profile of Gammatones makes them suitable to model temporal masking in auditory filterbanks~\cite{Fastl2007}.
Yet, the introductory paper on Gammatone wavelets \cite{Venkitaraman2014} does not provide a formula for deducing $\sigma$ from the specification of a quality factor $Q$.
In this appendix, we provide a rationale for choosing the topmost center frequency $\xi$ of a Gammatone wavelet filter bank in a discrete-time setting.
Then, we relate the bandwidth parameter $\sigma$ to the choice of a quality factor Q.
We express both $\xi$ and $\sigma$ in radians, so that converting this value to Hertz would incur a multiplication by $2\pi/f_s$ where $f_s$ is the sampling frequency of the audio signal at hand.

@@

The complex-valued Gammatone wavelet is a modification of the real-valued
Gammatone auditory filter, originated in auditory physiology. The
Gammatone auditory filter of center frequency $\xi$ is defined
as a gamma distribution of order $N\in\mathbb{N}^{*}$ and bandwidth
$\sigma$ modulated by a sine wave, that is,
\[
t^{N-1}\exp(-2\pi\sigma t)\cos(2\pi \xi t).
\]
For a fixed $\sigma$, the integer $N$ controls the relative shape
of the envelope, becoming less skewed as $N$ increases. Psychoacoustical
experiments have shown that, for $N=4$, the Gammatone function provides
a valid approximation of the basilar membrane response in the mammalian
cochlea \cite{Flanagan1960,Patterson1976,Lyon2010}. In particular,
it is asymmetric both in the time domain and in the Fourier domain,
which allows to reproduce the asymmetry of temporal masking as well
as the asymmetry of spectral masking \cite{Fastl2007}. It is thus
used in computational models for auditory physiology \cite{Pressnitzer2005}.
However, it does not comply with the Grossman-Morlet admissibility
condition, because it has a non-negligible average. In addition, because
the Gammatone auditory filter takes real values in the time domain,
its Fourier transform satisfies Hermitian symmetry, which implies
that it does not belong to the space $H^{2}$ of analytic functions.
More generally, there are no real-valued functions in $H^{2}$ \cite{Grossmann1984}.

@@

With the aim of building a pseudo-analytic admissible Gammatone wavelet,
\cite{Venkitaraman2014} have modified the definition of the Gammatone
auditory filter, by replacing the real-valued sine wave $\cos(2\pi t)$
by its analytic part $\exp(2\pi\mathrm{\mathrm{i}}t)$ and by taking
the first derivative of the gamma distribution, thus ensuring null
mean. The definition of the Gammatone wavelet becomes
\[
\boldsymbol{\psi}(t)=\left(2\pi(\mathrm{i}-\sigma)t^{N-1}+(N-1)t^{N-2}\right)\exp(-2\pi\sigma t)\exp(2\pi\mathrm{i}\xi t)
\]
in the time domain, and
\[
\boldsymbol{\widehat{\psi}}(\omega)=\dfrac{\mathrm{i}\omega\times(N-1)!}{(\sigma+\mathrm{i}(\omega-\xi))^{N}}
\]
in the Fourier domain. Besides its biological plausibility, the Gammatone
wavelet enjoys a near-optimal time-frequency localization with respect
to the Heisenberg uncertainty principle. Furthermore, this time-frequency
localization tends to optimality as $N$ approaches infinity, because
the limit $N\rightarrow+\infty$ yields a Gabor wavelet \cite{Cohen1995}.
Last but not least, the Gammatone wavelet transform of finite order
$N$ is causal, as opposed to the Morlet wavelet transform, which
makes it better suited to real-time applications. From an evolutionary
point of view, it has been argued that the Gammatone reaches a practical
compromise between time-frequency localization and causality constraints \cite{Venkitaraman2014}.
